Mokken scale analysis of the perioperative thirst discomfort scale

### 2020-08-12 manuscript/manuscript.Rmd
pdf

---
title: A Template for Two or One-Column Vignettes

@@

# Optional: line of arbitrary text with additional information.
# Could be used, for example, to mention the bibliographic info in a post-print.
# If not specified, defaults to "This version was compiled on \today"
#date_subtitle: Published in *Journal of Statistical Software*, 2018

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Author and Author

# Place eg a DOI URL or CRAN Package URL here
doi_footer: "https://cran.r-project.org/package=YourPackage"

# Abstract
abstract: |
  Your abstract will be typeset here, and used by default a visually distinctive font.
  An abstract should explain to the general reader the major contributions of the article.

# Optional: Acknowledgements
acknowledgements: |
  This template package builds upon, and extends, the work of the excellent
  [rticles](https://cran.r-project.org/package=rticles) package, and both packages rely on the
  [PNAS LaTeX](http://www.pnas.org/site/authors/latex.xhtml) macros. Both these sources are
  gratefully acknowledged as this work would not have been possible without them.  Our extensions
  are under the same respective licensing term
  ([GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) and
  [LPPL (>= 1.3)](https://www.latex-project.org/lppl/)).

# Optional: One or more keywords
keywords:
  - one
  - two
  - optional
  - keywords
  - here

# Paper size for the document, values of letter and a4
papersize: letter

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
#one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
#one_sided: true

# Optional: Enable section numbering, default is unnumbered
#numbersections: true

# Optional: Specify the depth of section number, default is 5
#secnumdepth: 5

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

@@

# Optional: Enable a 'Draft' watermark on the document
#watermark: true

@@

# Produce a pinp document
output: pinp::pinp

---

@@

## Abstract

@@

## Introduction

@@

## Methods

### Study Design

@@

### Participants

#### Inclusion criteria

@@

#### Exclusion criteria

@@

### Data collection

@@

#### Measures

@@

### Statistical analysis

@@

## Results

@@

### Mokken scale analysis

@@

#### Invariant item ordering

@@

#### Accuracy and reliability

@@

#### Item difficulty hierarchy

@@

```{r }
drake::readd(item_hierarchy_df_ptds5) %>% 
  dplyr::select(item) 
```

### Construct and convergent validity

The median PTDS-5 score was `r ptds5`. The distribution of PTDS-5 scores was right-skewed (Figure \@ref(fig:distribution)). There was a positive correlation between PTDS-5 scores and thirst intensity (Figure \@ref(fig:intensity)), as well as the global thirst discomfort rating (Figure \@ref(fig:globalthirst)). The associations between PTDS-5 scores with food and fluid duration is displayed in Figure \@ref(fig:fasting). Longer duration of fasting was not linearly associated with PTDS-5 scores. Likewise, the correlation between pain and PTDS-5 scores was not significant (Spearman's rho = `r round(drake::readd(pain_corr)$rho, 2)`; 95% CI = `r round(drake::readd(pain_corr)$CI_low, 2)` to `r round(drake::readd(pain_corr)$CI_high, 2)`).

## Discussion

Results indicate that the PTDS-5 is a reliable and precise way to order patients according to their thirst discomfort in the *pre-procedure* period. In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms of thirst discomfort in the same way. As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are expriencing worse thirst discomfort. The hierarchical pattern of responses to items in the Mokken scale is interpretable in terms of respondents more easily endorsing items related to a general desire to drink water through to first endorsing more specific symptoms related to dryness of the mouth and lips prior to endorsing discomfort associated with the sensations of 'thick' saliva and 'thick' tongue. In more practical terms, a patient who reports feeling bothered by a sensation that saliva is thick, could be interpreted as experiencing worse thirst discomfort than a patient who did not endorse any of the items lower down on the item difficulty hierarchy (i.e. 'my lips are dry' and 'my mouth is dry').

@@

### Limitations

@@

@@

```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
# knitr::include_graphics("figures/distribution.png")
drake::readd(distribution_plot)
```

```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
# knitr::include_graphics("figures/global-thirst.png")
library(dplyr)
drake::readd(global_thirst_plot)
```

>>>>

---
title: Mokken scale analysis of the perioperative thirst discomfort scale

@@

date_subtitle: "Blinded for peer review"
# Could be used, for example, to mention the bibliographic info in a post-print.
# If not specified, defaults to "This version was compiled on \today"

# For footer text  TODO(fold into template, allow free form two-authors)
lead_author_surname: Conway et al.

# Place eg a DOI URL or CRAN Package URL here
doi_footer: "Data and statistical code is available [here](https://github.com/awconway/ptds)"

# Abstract
abstract: |
   *Introduction:* Symptoms of thirst discomfort are common in patients who are required to undergo long periods of fasting prior to medical and surgical procedures. The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures.
   *Methods:* A prospective observational study was conducted. Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included. Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale. Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity ratings.
   *Results:* 
   *Conclusion:* The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort. Nurses may use this simple scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.

# Optional: One or more keywords
keywords:
  - fasting
  - anesthesia
  - thirst discomfort
  - patient-reported outcomes
  - scale validation
  - mokken scale analysis

@@

@@

# Produce a pinp document
output: pinp::pinp
params: 
  word: false
---

@@

```{r eval=params$word, results='asis'}
cat(
"
## Abstract

@@

"
)
```

# Introduction

@@

# Methods

## Study Design

@@

## Participants

### Inclusion criteria

@@

### Exclusion criteria

@@

## Data collection

@@

### Measures

@@

## Statistical analysis

@@

# Results

@@

## Mokken scale analysis

@@

### Invariant item ordering

@@

### Accuracy and reliability

@@

### Item difficulty hierarchy

@@

```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
drake::readd(item_hierarchy_df_ptds5) %>% 
  kableExtra::kable(booktabs=T)
```

## Construct and convergent validity

The median PTDS-5 score was `r ptds5`. The distribution of PTDS-5 scores was right-skewed (Figure \@ref(fig:distribution)).

```{r distribution, fig.cap = "Distribution of PTDS-5 scores (higher scores indicate worse thirst discomfort)"}
# knitr::include_graphics("figures/distribution.png")
drake::readd(distribution_plot)
```

There was a positive correlation between PTDS-5 scores and thirst intensity (Figure \@ref(fig:intensity)), as well as the global thirst discomfort rating (Figure \@ref(fig:globalthirst)). The associations between PTDS-5 scores with food and fluid duration is displayed in Figure \@ref(fig:fasting).

```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating", eval=params}
# knitr::include_graphics("figures/global-thirst.png")
library(dplyr)
drake::readd(global_thirst_plot)
```

Longer duration of fasting was not linearly associated with PTDS-5 scores. Likewise, the correlation between pain and PTDS-5 scores was not significant (Spearman's rho = `r round(drake::readd(pain_corr)$rho, 2)`; 95% CI = `r round(drake::readd(pain_corr)$CI_low, 2)` to `r round(drake::readd(pain_corr)$CI_high, 2)`).

# Discussion

Results indicate that the PTDS-5 is a reliable and precise way to order patients according to their thirst discomfort in the *pre-procedure* period. In addition, an acceptable accuracy for invariant item ordering was identified, meaning patients are likely to rank the PTDS-5 symptoms of thirst discomfort in the same way. As such, the item difficulty hierarchy, presented in Table \@ref(tab:tab2) can be used as a guide to easily determine which patients are experiencing worse thirst discomfort. The hierarchical pattern of responses to items in the Mokken scale is interpretable in terms of respondents more easily endorsing items related to a general desire to drink water through to first endorsing more specific symptoms related to dryness of the mouth and lips prior to endorsing discomfort associated with the sensations of 'thick' saliva and 'thick' tongue. In more practical terms, a patient who reports feeling bothered by a sensation that saliva is thick, could be interpreted as experiencing worse thirst discomfort than a patient who did not endorse any of the items lower down on the item difficulty hierarchy (i.e. 'my lips are dry' and 'my mouth is dry').

@@

## Limitations

@@

## Conclusion

@@
